---
title: "Wrangle data for suplementary files with all DEG from comparisons and groups"
author: "Mauricio Losilla"
date: "March 16, 2019"
output: html_document
---

Load packages and set wd
```{r setup}
# Load required libraries
library(knitr) #only needed to knit Rmarkdown file
library(dplyr)
library(xlsx)

session.Info <- sessionInfo()

#set the wd. This is where the files with the read counts and the output folder are
knitr::opts_knit$set(root.dir = "/home/mau/Dropbox (MSU Efish Lab)/Mau/Research/Differential expression/Analyses/05_Differential_Gene_Expression/R") #set this one if knitting the R markdown file
setwd("/home/mau/Dropbox (MSU Efish Lab)/Mau/Research/Differential expression/Analyses/05_Differential_Gene_Expression/R") #set this one if using the console, or if changing the wd for only one chunk of code in the .Rmd file

#the following will allow par changes for specific plots
#knitr::opts_knit$set(global.par = TRUE)
```


Import Data: geneID-to-name dictionary
```{r}

#Import the file that lists Pking GeneIDs, gene names & symbols
Pkings.gene.name.type <- read.table("Dic.PkingEntrezGeneID-to-name_symbol_type.txt", header = TRUE, row.names = NULL, sep = "\t", col.names = c("Pking_Entrez_geneID", "Pking_gene_name", "Pking_gene_symbol", "type_of_gene"), colClasses = c("character", "character", "character", "character"), quote = "")

```


Import Data from Trinity: DEG from each comparison
```{r}
#make a vector with the names of the files to import
DEG.files <- list.files(path = "../genes_edgeR/", pattern = "RSEM.gene.counts.matrix.*results.P0.001_C2.DE.subset", full.names = TRUE)

#extract the names of each comparison
comparisons <- lapply(DEG.files, function(x) {sub(".*matrix.", "", x) %>% sub(".edge.*", "", .)})

#Import the DEG for each comparison
DEG.all <- lapply(DEG.files, function(x) as.data.frame(read.table(x, header = TRUE, row.names = NULL, sep = "\t", quote = "", stringsAsFactors =FALSE)))
DEG.all <- lapply(DEG.all, function(x) {names(x)[names(x) == 'row.names'] <- "Pking_Entrez_geneID"; x})

#name each data frame with the respective comparison
names(DEG.all) <- comparisons
```


Import Data: DEG from each group (cluster)
```{r}
#make a vector with the names of the files to import
groups.files <- list.files(path = "../Custom_clusters/clusters_logTMM/", pattern = "enriched_in*", full.names = TRUE)

#extract the names of each phenotype
phenotypes <- lapply(groups.files, function(x) {sub(".*_in_", "", x) %>% sub("_samples__.*", "", .)})

#Import the DEG for each group
groups.all <- lapply(groups.files, function(x) as.data.frame(read.table(x, header = TRUE, row.names = NULL, sep = "\t", quote = "", stringsAsFactors =FALSE)))
groups.all <- lapply(groups.all, function(x) {names(x)[names(x) == 'row.names'] <- "Pking_Entrez_geneID"; x})

#name each data frame with the respective phenotype
names(groups.all) <- phenotypes
```


The upregulated genes in the groups are listed per phenotype. Merge the phenotypes into their EOD feature
```{r}
##add a column with the phenotype
groups.all.phen <- groups.all
groups.all.phen <- lapply(names(groups.all.phen), function(x) {groups.all.phen[[x]]["enriched phenotype"] <- x; groups.all.phen[[x]]})
names(groups.all.phen) <- phenotypes


## We are ready to merge the phenotypes per feature

#make vector with the features
#features <- c("complexity", "duration", "polarity")

#create an empty list to hold the results
groups.feature <- list()

#merge the phenotypes per feature
groups.feature[["duration"]] <- groups.all.phen$short_EOD %>% bind_rows(., groups.all.phen$long_EOD)
groups.feature[["polarity"]] <- groups.all.phen$small_penetrations %>% bind_rows(., groups.all.phen$large_penetrations)
groups.feature[["complexity"]] <- groups.all.phen$biphasic %>% bind_rows(., groups.all.phen$triphasic)

```


Add gene name, symbol and type; to 1) the DEG of each comparison, and 2) the DEG of each feature (groups)
```{r}
## 1) to the DEG of each comparison
#add the gene name, symbol and type of gene from the dictionary, and reorder columns
DEG.all.complete <- lapply(DEG.all, function(x) {left_join(x, Pkings.gene.name.type, by = "Pking_Entrez_geneID") %>% select(c(1, 19:21, 2:18))})

## 2) to the DEG of each feature (groups)
#add the gene name, symbol and type of gene from the dictionary, and reorder columns
groups.feature.complete <- lapply(groups.feature, function(x) {left_join(x, Pkings.gene.name.type, by = "Pking_Entrez_geneID") %>% select(c(1, 14:16, 13, 2:12))})
```


save to Xcel files, and save the workspace
```{r}
## save the comparisons file

#create empty workbook
wb <- createWorkbook()
saveWorkbook(wb, "Suppl_file_DEG_all_comparisons.xlsx")

#save the DEG from each comparison, in a tab with the name of the comparison
lapply(names(DEG.all.complete), function(x) write.xlsx(DEG.all.complete[[x]], "Suppl_file_DEG_all_comparisons.xlsx", sheetName = x, col.names = TRUE, row.names = FALSE, append = TRUE))


## save the features (groups) file

#create empty workbook
wb <- createWorkbook()
saveWorkbook(wb, "Suppl_file_DEG_groups.xlsx")

#save the DEG from each feature, in a tab with the name of the feature
lapply(names(groups.feature.complete), function(x) write.xlsx(groups.feature.complete[[x]], "Suppl_file_DEG_groups.xlsx", sheetName = x, col.names = TRUE, row.names = FALSE, append = TRUE))


#save the workspace
save.image(file = "generate_suppl_files_DEG_comparisons_and_groups.RData")
```
